---
title: "When Was the Last Day We Hit 20c?"
author: Nicholas Tierney
date: '2020-09-07'
slug: last-day-20c
categories:
  - rstats
tags:
  - rstats
  - bomrang
output: hugodown::md_document
---

```{r setup, include = FALSE}
options(cli.width = 70)  # For tidyverse loading messages
knitr::opts_chunk$set(
  tidy.opts = list(width.cutoff = 70),  # For code
  width = 70,
  collapse = TRUE, 
  comment = "#>", 
  fig.width = 7, 
  fig.align = 'center',
  fig.asp = 0.618, # 1 / phi
  fig.retina = 2,
  out.width = "700px"
)
```

It's starting to get warmer here in Melbourne. Today was the first day of over 20 degrees Celsius we've had in...well how long? It feels like a while. We hit this middle part of the year and we don't get over 20C for a while. But I'd like to know the last time it got over 20 degrees Celsius. How do we do that? With the [`bomrang`](https://github.com/ropensci/bomrang) package by [Adam Sparks](https://github.com/adamhsparks)!

First, you'll need the latest version of `bomrang` - you should be on version 0.7.0.9000 or greater:

```{r install}
# remotes::install_github("ropensci/bomrang")
suppressPackageStartupMessages(library(bomrang))
packageVersion("bomrang")
```

Next, we sweep for stations nearby Melbourne. I know there are packages (like [geocoder]()) that help get the coordinates, but when I just need one coordinate I usually just copy it from google maps. 

```{r melb-stations}
melb_stations <- sweep_for_stations(latlon = c(-37.800372,
                                                        144.996162))

melb_stations
```

This then returns information on the number of stations in Melbourne and their location.

We take the first station and get the max temperature.

```{r melbourne-weather}
melbourne_weather <- get_historical(stationid = melb_stations$site[1],
                                    type = "max")

melbourne_weather
```

We can convert this into a date format with one of my favourite R functions of all time, `ISOdate()` (also `ISOdatetime()` is great):

```{r melb-weather-date}
melb_weather_date <- melbourne_weather %>% 
  mutate(date = ISOdate(year, month, day))

melb_weather_date
```

Now we can look at when it last hit 20C for the past little while:

```{r plot-temps}
suppressPackageStartupMessages(library(ggplot2))

ggplot(melb_weather_date,
       aes(x = date,
           y = max_temperature)) + 
  geom_line() + 
  geom_hline(yintercept = 20, colour = "salmon")

```

Let's just focus on 2020:

```{r weather-2020}
suppressPackageStartupMessages(library(dplyr))
melb_weather_2020 <- melb_weather_date %>% 
  filter(year == 2020)

melb_2020_plot <- 
ggplot(melb_weather_2020,
       aes(x = date,
           y = max_temperature)) + 
  geom_line() + 
  geom_hline(yintercept = 20, colour = "salmon")

melb_2020_plot
```

So now rather than squinting at a graph, I want to return the last times it was over 20C.

I was tempted to try something like this:

```{r try-filter}
melb_weather_2020 %>% 
  filter(max_temperature > 20) %>% 
  ggplot(aes(x = date,
           y = max_temperature)) + 
  geom_line() + 
  geom_hline(yintercept = 20, colour = "salmon")
```

But this doesn't realllly help me. We just see a bit gap. I still need to squint. Let's take a look at the plot again:

```{r maybe-plot}
melb_2020_plot
```

I want to be able to identify when it stopped being over 20, and then when it started again. 

Buckle up, we're going into the realm of RLE - Run Length Encodings.

[Earo Wang](https://github.com/earowang) first showed this to me I first arrived at Monash, (where it was used for [`miss_var_run`](http://naniar.njtierney.com/reference/miss_var_run.html)). 

`rle` counts the length of a "run" of a vector. Here's an example to explain from the help file:

```{r rle}
z <- c(TRUE, TRUE, FALSE, FALSE, TRUE, FALSE, TRUE, TRUE, TRUE)
rle(z)
```

So this tells us that we have 2 runs of the same thing, then 2 more, then 1, 1, and 3. Another way I think about it is is like a simplified/summary of storing a number. Here's another brief example to help get this solidified:

```{r rle-again}
rle(1:10)
rle(c(1,2,2,3,3,3))
rle(c(1,1,1,2,2,3))
```

So, we can use `rle` to calculate the run of the times that temperature was below 20. We can calculate a new column that is TRUE when temperature is below, and FALSE otherwise: 

```{r melb-rle-mutate}
melb_weather_2020 %>% 
  mutate(below_20 = max_temperature < 20) 
```

then `pull` it out

```{r melb-rle-pull}
melb_weather_2020 %>% 
  mutate(below_20 = max_temperature < 20) %>% 
  pull(below_20) 
```

And then use `rle` on the vector of TRUE/FALSE

```{r melb-rle-calc}
melb_rle <- melb_weather_2020 %>% 
  mutate(below_20 = max_temperature < 20) %>% 
  pull(below_20) %>% 
  rle()

melb_rle
```

This provides two named vectors, `lengths` and `values`. Lengths is the number of times a corresponding value repeats.

We can then identify what the largest gap was by taking the max of `lengths`:

```{r max-rle}
# what was the largest gap?
what_gap <- max(melb_rle$lengths)

what_gap
```

Neat! So we now know that the period between the start and the end spanned `r what_gap` days. 

So, _when_ was that?

This part involves a few steps, let me break it down.

We want to find that date it was last over 20C - which is the day before it had that 111 day stretch. We can get this by summing up all the numbers of run lengths before the 111 day stretch. 

First we calculate which position has the max with `which.max`. I like to print the vector with it for my own sanity's sake:

```{r which-max-length}
which.max(melb_rle$lengths)
melb_rle$lengths
```

So it is the `r which.max(melb_rle$lengths)`th one.

But we want the _second_ last one - the day before 111, so we subtract 1.

```{r last-day-position}
last_day_position <- which.max(melb_rle$lengths) - 1
last_day_position
```

We then want 1 through to that number so we can get all the numbers out, using `seq_len`:

```{r seq-len}
seq_last_day_position <- seq_len(last_day_position)
seq_last_day_position
```

We then subset this lengths like so, and sum them up to get the number of days **before** it went below 20C:

```{r rle-before-20}
melb_rle$lengths[seq_last_day_position]
n_days_before_20c <- sum(melb_rle$lengths[seq_last_day_position])
n_days_before_20c
```

Now...what date is that?

We can use `lubridate` to calculate it like so, by saying, the number of days from the start of 2020:

```{r when-was-that}
library(lubridate)
# when was that?
last_day_of_20c <- ymd("2020-01-01") + days(n_days_before_20c)
last_day_of_20c
```

Awesome, now when did it finish?

```{r when-did-it-end}
# when did it end?
first_day_of_20c <- last_day_of_20c + days(what_gap)
first_day_of_20c
```

And just to come full circle, let's check the difference

```{r difference}
first_day_of_20c - last_day_of_20c
```

whew! That was a lot of work. I almost forgot what I was trying to answer so let's put this in a header now

# When was the last time in 2020 Melbourne was above 20 Celsius?

* The last day of over 20C was `r last_day_of_20c`
* The first day of over 20C was `r first_day_of_20c`
* Which means we had `r first_day_of_20c - last_day_of_20c` days where Melbourne was below 20C.

# Post Script

In retrospect there is more to add to this plot, and probably other simpler ways to calculate this, I'm keen to hear your thoughts below.

I suspect this sort of function will come in handy in the future as we start to compare COVID19 days, to help us answer: "when was the last day of ... cases".

~~I'm trying to get this blog post out in one sitting so I'll skip over writing a function for this, this time~~ Here is a function and an example below of this, which might be useful:

```{r functionalisationing}
n_days_below <- function(x, below){
  rle_below <- rle(x < below)
  max(rle_below$lengths)
}

last_day_below <- function(x, below){
  rle_below <- rle(x < below)
  last_day_position <- which.max(rle_below$lengths) - 1
  seq_last_day_position <- seq_len(last_day_position)
  n_days_before_below <- sum(rle_below$lengths[seq_last_day_position])
  n_days_before_below
}

summary_days_below <- function(x, below, origin){
  
  vec_n_days_below <- n_days_below(x, below)
  vec_last_day_below <- n_days_below(x, below)
  
  last_day_of_below <- lubridate::ymd(origin) + lubridate::days(vec_last_day_below)
  first_day_of_below <- last_day_of_below + vec_n_days_below
  
  list(last_date_below = last_day_of_below,
       first_date_above = first_day_of_below,
       time_between = first_day_of_below - last_day_of_below)
  
}

n_days_below(melb_weather_2020$max_temperature, 20)
last_day_below(melb_weather_2020$max_temperature, 20)
summary_days_below(melb_weather_2020$max_temperature, 20, "2020-01-01")

```


# A side note on the history of `bomrang`

I take great joy in seeing `bomrang` and using it. I first posted [an issue](https://github.com/ropensci/unconf16/issues/24) about it in 2016 at the [rOpenSci unconf]() which was held in the USA, but [Miles](https://milesmcbain.xyz/), [Jessie](https://acems.org.au/our-people/jessie-roberts), and I (And also [Alex Simmons](https://scholar.google.com.au/citations?user=plaGy54AAAAJ&hl=en)) participated remotely, with [Bob Rudis](https://rud.is/b/), [Brooke Anderson](https://github.com/geanders), [Maëlle Salmon](https://masalmon.eu/) and a few folks. My memory is a bit hazy because in order to keep pace with the USA we had to stay up all night, and I don't really remember the days before or after it. It got somewhere, but I didn't really know what we were doing, I just thought, "this should be a thing, why can't I get Australian weather data from R?", but had no idea what I was doing with web scraping, and knew nothing about the weather other than what I'd seen on the news and that lightning can indeed strike the same place twice. So I suggested the idea, but everyone else provided the know how. It was hard, because it turned out the BoM weather bureau API didn't exist and required a bit of domain expertise

Next, it was [picked up by much more experienced team](https://github.com/ropensci/auunconf/issues/6) at the first [Australian rOpenSci Ozunconf](https://github.com/ropensci/auunconf) (then, au-unconf), in 2016 in Brisbane. Happily they knew what they were doing, and has experience with Australian weather data. The project grew from there, being headed by [Adam Sparks](https://adamhsparks.com/), initially it was called `bomr`, but [Di Cook](http://dicook.org/) suggested, "bomrang", the name stuck, and Adam submitted it to rOpenSci for package review, and the package became more mature and in it's current state of glory. 
