---
title: Quickly Assessing Colour Palettes
author: Nicholas Tierney
date: '2020-10-15'
slug: assess-colour
categories:
  - data visualisation
  - colour palette
  - rstats
tags:
  - data visualisation
  - colourblind
  - colorblind
  - rstats
output: hugodown::md_document
---

```{r setup, include = FALSE}
options(cli.width = 70)  # For tidyverse loading messages
knitr::opts_chunk$set(
  tidy.opts = list(width.cutoff = 70),  # For code
  width = 70,
  collapse = TRUE, 
  comment = "#>", 
  fig.width = 7, 
  fig.align = 'center',
  fig.asp = 0.618, # 1 / phi
  fig.retina = 2,
  out.width = "700px"
)
```

You want to use a nice colour palette but you're not sure if it's colour blind friendly? Here are some quick ways to check this in #rstats.

# Using prismatic

The [prismatic](https://github.com/EmilHvitfeldt/prismatic) package by [Emil Hvitfeldt](https://github.com/EmilHvitfeldt) provides some nice approaches to this with the function `check_color_blindness()`. You provide a vector of colour codes, and it simulates how they appear for people with different types of colourblindness.

For example, we can check the "Cold" palette from the `qualitative_chl` colour palette from [colorspace](https://colorspace.r-forge.r-project.org/index.html) by [Achim Zeileis](https://eeecon.uibk.ac.at/~zeileis/) and co, like so:

```{r example}
library(colorspace)
hcl_cold <- 
qualitative_hcl(n = 3,
                palette = "Cold")

library(prismatic)
check_color_blindness(hcl_cold)
```

But what if you want to check a few palettes in quicker succession? I used a function like this to help. First, getting a handle on what qualitative colour palettes are available:

```{r hcl-avail}
hcl_pals_q <- hcl_palettes(type = "qualitative")
hcl_pals_q
```

Then writing a function to pick three colours. I also needed a specific alpha level of 0.67, but that was just me. I called it `qual_cvd` because it assesses qualitative colour palettes for colour vision deficiency (CVD).

```{r fun-check-cvd}
library(magrittr)
qual_cvd <- function(palette = "Pastel1", alpha = 0.67){
  qualitative_hcl(n = 3,
                  alpha = alpha,
                  palette = palette)  %>% 
    check_color_blindness()
}
```

Then we can demonstrate it like so:

```{r demo-check-cvd}
qual_cvd("Dark 2")
qual_cvd("Dark 3")
```

Looks like Dark 3 is a good one!

# Using colorspace

The `colorspace` package provides some really nice functions for exploring CVD.

One thing I really love here is the `demoplot` function, which provides an example of what a plot might look like with a given colour palette. For example:

```{r qual-demoplot}
qualitative_hcl(n = 3,
                alpha = 0.67,
                palette = "Cold") %>% 
  demoplot(type = "map")
```

There are many other plot types you can choose from, ("map", "heatmap", "scatter", "spine", "bar", "pie", "perspective", "mosaic", and "lines").

You can combine your returned colour hex codes with the `protan()`, `deutan()`, and `tritan()` functions to simulate a particular type of CVD. For example, convert the colours into protanopia:

```{r protan-show}
qualitative_hcl(n = 3,
                alpha = 0.67,
                palette = "Cold") %>% 
  protan()
```


Let's wrap this up into a function to return the colour palettes as a list.

```{r qual-demoplot-fun}
qual_return_cvd <- function(palette = "Cold", 
                            alpha = 0.67){
  pals <- qualitative_hcl(n = 3,
                          alpha = alpha,
                          palette = palette)
  pals_protan <- protan(pals)
  pals_deutan <- deutan(pals)
  pals_tritan <- tritan(pals)
  
  return(list(
    protan = pals_protan,
    deutan = pals_deutan,
    tritan = pals_tritan
    ))
}
qual_return_cvd()
```

Why return a list? Then we can use `purrr::walk()` to iterate over it (`purrr::map()` also works, but we just want the side effect, the plot, so walk is fine here).

```{r walk-cvd}
qual_return_cvd() %>%  purrr::walk(demoplot)
```

Hmm, Ok, so not the best, let's try dark 3 again

```{r walk-cvd-dark}
qual_return_cvd(palette = "Dark 3") %>%  purrr::walk(demoplot)
```

We get much better separation between the colours, so I think I'd go with "Dark 3".

What is really lovely about the `demoplot` function is that you can get a quick sense of what your selected palette might look like for a given type of plot, without needing to go to the hassle of putting it through your data. This quick iteration is really key, I think.

# End

There is _always_ more to talk about when it comes to colours, and this is just a short post on the topic - I've left a lot out of it, otherwise it wouldn't ever get finished! I do have an upcoming in depth blog post series I've been working on that explains the details of what colour blindness is, but it has been a work in progress for about a year. So, I figured I'd rather get this post out quickly and keep it brief.

If you'd like to learn more about colour blindness in graphics, I gave a talk at Monash Data Fluency about this earlier this year, entitled, "The Use of Colour in Graphics" [slides and materials are available here](https://github.com/njtierney/monash-colour-in-graphics), which also has some nice resources provided on where you can learn more.
