---
title: 'Tidyverse Case Study: Exploring XKCD comics'
author: ''
draft: true
date: '2017-11-09'
slug: tidyverse-xkcd
categories:
  - rstats
  - rbloggers
tags: []
output: hugodown::hugo_document
---

I'm a huuuge fan of XCKD.

Randall Munroe has done some awesome stuff, including:

- Writing about a submarine, landing him a job in NASA
- asking people about their colour preferences.

XKCD is also a cool name - you need to say each of the letters, there's no real shortcut word for it.

Once again, reading Joe's post about Data packages, I came across the XKCD package - which is really really awesome.

It provides a whole bunch of info about the XKCD comics.

Randall now writes the comics full time, along with his other really awesome books - 

After going through the Billboard data, I am wondering if I can explore the rise of XKCD - perhaps we can track when the comics become regular?

```{r}
library(XKCDdata)
library(tidyverse)
library(future)
library(furrr)
library(tictoc)

null_comic <- get_comic(1) %>% 
  mutate(across(everything(), function(x) x <- NA))

possibly_get_comic <- possibly(get_comic, otherwise = null_comic)

plan(multisession, workers = 6)
xkcd_data <- furrr::future_map_dfr(1:2374, possibly_get_comic)
```


```{r}
xkcd_data %>%
  slice_head(n = 1) %>% 
  pivot_longer(cols = everything(),
               names_to = "vars",
               values_to = "vals",
               values_transform = as.character)

library(lubridate)

xkcd_data_cut <- xkcd_data %>% 
  select(num, year, month, day, title, alt, transcript) %>% 
  mutate(date = as_date(ISOdate(year,month,day)),
         .after = num)  %>% 
  select(-c(year, month, day)) %>% 
  mutate(wday = wday(date, label = TRUE),
         .after = date) %>% 
  padr::pad()

View(xkcd_data_cut)

xkcd_data_cut %>% 
  ggplot(aes(x = date,
             y = num)) + 
  geom_point()

```

# Future Work

I'm also a big fan of the webcomics, Sunday Morning Breakfast Cereal, Perry Bible Fellowship, and Questionable Content. Perhaps these can make some nice toy R packages in the future.
