---
title: 'Just Quickly: Fizzbuzz'
author: ''
date: '2019-07-24'
draft: true
slug: jq-fizzbuzz
categories:
  - rstats
tags:
  - rstats
  - data science
output: hugodown::hugo_document
---

One of my favourite YouTubers, [Tom Scott](https://www.youtube.com/watch?v=QPZ0pIK_wsc) has a video about `fizzbuzz`. This is a game I played at school in drama as a "warm up" game. The rule is this:


In his video, Tom says, "have a go now if you like", and I thought, "sure".

So, here is my approach to how I would do fizzbuzz.


```{r}
# fizzbuzz

library(tidyverse)

df <- tibble(x = 1:100)

# checking modulo does what I think it does
3 %% 3
6 %% 3
9 %% 3
  
df %>%
  mutate(div_three = (x %% 3) == 0,
         div_five = (x %% 5) == 0,
         div_both = div_three & div_five,
         speak = case_when(
           div_three ~ "fizz",
           div_five ~ "buzz",
           div_both ~ "fizzbuzz",
           !div_three & !div_five & !div_both ~ paste0(x)
         ))

```


# Just what is going on here?

Load the tidyverse, and create a dataframe:

```{r create-df}
library(tidyverse)
df <- tibble(x = 1:100)
```

I like to create a dataframe to set this up. I work with data every day, and for me, this is my playing field. It feels comfortable.

Now, check if something is divisible by 3 or 6 or whatever:

```{r modulo}
# checking modulo does what I think it does
3 %% 3
6 %% 3
9 %% 3
```

If you haven't seen `%%` before - you are not alone. I first encountered this because Miles McBain saw it in an example somewhere, and _it is cryptic to google or search help for_. It is the `modulo` operator. This returns the remainder after you divide something. So, at the start, I'm checking that `%%` behaves like I think it does - I rarely use this function so it's good to check. Great, it returns 0 when dividible by a number.

Now, my plan is to create a logical column that is TRUE when dividible for three and five

```{r div-three}
df %>%
  mutate(div_three = (x %% 3) == 0,
         div_five = (x %% 5) == 0)
```


OK great, now to handle when three and five:

```{r div-both}
df %>%
  mutate(div_three = (x %% 3) == 0,
         div_five = (x %% 5) == 0,
         div_both = div_three & div_five)
```

And now to speak, using `case_when` - this is like a more generalised if statement. When the thing on the left hand side is true, output the thing on the right.

```{r nearly-done}
df %>%
  mutate(div_three = (x %% 3) == 0,
         div_five = (x %% 5) == 0,
         div_both = div_three & div_five,
         speak = case_when(
           div_three ~ "fizz",
           div_five ~ "buzz",
           div_both ~ "fizzbuzz"
         ))
```


OK, but now we've got some NAs, let's tell it to speak when
all are FALSE:

```{r eval = FALSE}
df %>%
  mutate(div_three = (x %% 3) == 0,
         div_five = (x %% 5) == 0,
         div_both = div_three & div_five,
         speak = case_when(
           div_three ~ "fizz",
           div_five ~ "buzz",
           div_both ~ "fizzbuzz",
           !div_three & !div_five & !div_both ~ glue::glue("{x}")
         ))
```

Ugh, it didn't like the glue string, let's use `paste0` instead:

```{r }
df %>%
  mutate(div_three = (x %% 3) == 0,
         div_five = (x %% 5) == 0,
         div_both = div_three & div_five,
         speak = case_when(
           div_three ~ "fizz",
           div_five ~ "buzz",
           div_both ~ "fizzbuzz",
           !div_three & !div_five & !div_both ~ paste0(x)
         ))
```

OK great.

# How to improve this?

I guess I would probably add a function like `is_div`, which would mean I could write it like this:

```{r}
is_div <- function(lhs, rhs) lhs %% rhs == 0

df %>%
  mutate(div_three = is_div(x, 3),
         div_five = is_div(x, 5),
         div_both = div_three & div_five,
         speak = case_when(
           div_three ~ "fizz",
           div_five ~ "buzz",
           div_both ~ "fizzbuzz",
           !div_three & !div_five & !div_both ~ paste0(x)
         ))
```


# What did Tom do?

Watching the rest of Tom's video, he uses a for loop to tackle the problem. This is a pretty fair approach, but I find that I don't really think about things in loops these days, I tend to think more about vectorisation.

Tom then ran some 


