---
title: Some Thoughts on Why Exploring Longitudinal Data is Hard
author: ''
date: '2019-07-25'
slug: longitudinal-data-is-hard
draft: true
categories:
  - rstats
  - longitudinal data
  - data visualisation
  - Statistics
tags:
  - rstats
output: hugodown::hugo_document
---

Back in 2019, I was thinking a lot about longitudinal data, as I was working on the [brolgar]() R package. I wrote down some thoughts on why exploring longitudinal data was hard, as I was finding it tricky. I've been trying to dust off some thoughts on the matter.

I think I've finally worked out how to express why it is hard to explore longitudinal data.

This is something I've been thinking about pretty intensely over the semester break.

I think that the reason is that we aren't explicitly trained in how to move from a data summary space back to the data space.

This means that when we have a dataset like the following:

```{r setup}
knitr::opts_chunk$set(eval = FALSE)
```

It is not intuitive to think about why choosing a given individual line is the one we select. 

Sure, it's interesting - it's at the bottom!

But why pick that one?

I think that the reason is we need to understand to move from a data summary space like the following, where each row is a series, or a key, or an individual, or whatever you want to call it.

```{r}

```

We then perform some way of identifying which of those individuals are interesting, keep them, then unpack those individuals back to their full data.

So, I think that the problem is that we don't think clearly in terms of stepping from the data space into a summary space, and then stepping back.

We need a tool to step back between the summary space and the data space - and that tool has existed in data bases for a long time.

It is the a join! i guess, specifically, a left join. The main join to know.

So what I need to do now is create a series of animations to show what I mean in these cases.

Something that collapses a series of lines down to a point, the points are then chosen, and unpacked back into lines.

I guess it looks like this (show code from Rstudioconf 2022)
