---
title: "Autoplot's, assemble! How to use `ggplot2::autoplot` to plot your new class"
author: ''
date: '2019-07-07'
slug: autoplot-ggplot2
draft: true
categories:
  - rstats
  - ggplot2
tags: []
output: hugodown::hugo_document
---

This blog post shows you how and why you might want to make your own `autoplot`. I wrote it because I was showing someone about `autoplot` a little while back and I couldn't find many full examples of defining your own autotplot, so I thought this blog post might be useful.

# What is autoplot? What is plot? What is life?

You might be familiar with the fact that you can use `plot` on an R object, like a data frame, and get a nice plot out:

```{r plot-mtcars}
plot(mtcars)
```

`plot` works with all kinds of things - for example you can plot a linear model and get a (or several!) nice plot(s) out:

```{r lm-fit-plot, out.width = "50%"}
lm_fit <- lm(mpg ~ ., mtcars)
plot(lm_fit)
```

These plots are quite different, yet they use the same function `plot`.

This is because behind the scenes, R uses S3 methods to handle different plotting methods. This is known as S3 method dispatch - so named because they came with version three of S (which was the precursor language to R). There's a great chapter on this in [Hadley Wickham's Advanced R](). I also once wrote a bit about S3 methods in [this arxiv article, "Simple S3 methods"]() - although in retrospect I'm not sure if my writing is as simple or useful as I thought.

## `autoplot` - `plot`, but for `ggplot`

[`ggplot2`]() has a similar plotting method, `autoplot`. A full featured example of `autoplot` is in the `fable` R package, from the tidyverts team, this example is [adapted from their website](https://fable.tidyverts.org/#example).

```{r}
library(tidyverse)
library(fable)
library(tsibble)
library(tsibbledata)
library(lubridate)

aus_retail_data <- aus_retail %>%
  filter(
    State %in% c("New South Wales", "Victoria"),
    Industry == "Department stores"
  )

aus_retail_model <- aus_retail_data %>% 
  model(
    ets = ETS(box_cox(Turnover, 0.3)),
    arima = ARIMA(log(Turnover)),
    snaive = SNAIVE(Turnover)
  )

aus_retail_model

aus_retail_forecast <- aus_retail_model %>% 
  forecast(h = "2 years")

aus_retail_forecast

aus_retail_forecast_plot <- aus_retail_forecast %>% 
  autoplot(filter(aus_retail, year(Month) > 2010), level = NULL)

aus_retail_forecast_plot
```

Now a lot is going on here, but the most important thing is this line of code here:

```{r}
#| eval: false
aus_retail_forecast_plot <- aus_retail_forecast %>% 
  autoplot(filter(aus_retail, year(Month) > 2010), level = NULL)

aus_retail_forecast_plot
```

Here they are using the `autoplot` function, along with some extra options, to create a ggplot object. You can modify this just like other ggplots as well - we could make it a minimal theme and move the legend position, see:

```{r}
aus_retail_forecast_plot +
  theme_minimal() +
  theme(legend.position = "bottom")
```

# Our turn - hi five data

So now let's make our own autoplot! First, let's create some motivating data. Let's say we have a dataset of hi fives, and the number of hi fives people of different age groups give. We have them binned into two year groups, from age 5 to 25

```{r}
five_to_25 <- seq(from = 5, to = 25, by = 5)
five_to_25
dat <- expand_grid(
  age_from = five_to_25,
  age_to = five_to_25
) %>% 
  mutate(
    n_hi_fives = rpois(25, 5)
  )

dat
```

## A digression: Hi fives are symmetric though...

I know that we are just imagining this dataset, BUT since hi-fives are symmetric - like, if I give 4 hi fives, 4 other people have received them, well, I think we should enforce that. A bit annoying, perhaps, but a quick google search [reveals a nice little bit of R code](https://stackoverflow.com/questions/18165320/creating-a-symmetric-matrix-in-r). So let's start by making a little matrix, then we'll make it symmetric.

```{r}
hi_five_mat <- matrix(data = rpois(25, 5),
                      nrow = 5,
                      ncol = 5,
                      dimnames = list(five_to_25, five_to_25))
hi_five_mat

hi_five_mat[lower.tri(hi_five_mat)] <- t(hi_five_mat)[lower.tri(hi_five_mat)]

hi_five_mat
```

And then, because we're really cool, we're going to turn these into functions, because that's fun.

```{r}
make_hi_fives <- function(rate = 10){
  five_to_25 <- seq(from = 5, to = 25, by = 5)
  mat <- matrix(data = rpois(25, lambda = rate),
                nrow = 5,
                ncol = 5,
                dimnames = list(five_to_25, five_to_25))
  # make a stronger diagonal pattern
  diag(mat) <- rpois(5, 20)
  mat
}

make_symmetric <- function(x){
  x[lower.tri(x)] <- t(x)[lower.tri(x)]
  x
}

hi_fives <- make_hi_fives()
# not symmetric!
hi_fives

# symmetric!
make_symmetric(hi_fives)
```

OK and then let's make this into a dataframe, because that's where we live.

```{r}
# data frame?
make_symmetric(hi_fives) %>% 
  as_tibble(rownames = "age_to") %>% 
  pivot_longer(
    cols = -age_to,
    names_to = "age_from",
    values_to = "n_hi_five"
  )

make_symmetric_hi_fives <- function(rate = 5){
  hi_fives <- make_hi_fives(rate = rate)
  make_symmetric(hi_fives) 
}

hi_fives_to_df <- function(df){
  df %>% 
  as_tibble(rownames = "age_to") %>% 
  pivot_longer(
    cols = -age_to,
    names_to = "age_from",
    values_to = "n_hi_five"
  )
}

hi_five_df <- make_symmetric_hi_fives() %>% 
  hi_fives_to_df()

hi_five_df
```

OK so...where were we? Back to making plots!

# Making plots of hi fives

In this case we'll just make a plot of some hi five data like so

```{r}
library(colorspace)
ggplot(hi_five_df,
       aes(x = age_to,
           y = age_from,
           fill = n_hi_five)) + 
  geom_raster() +
  scale_fill_continuous_sequential("Purples")
```

So what we're seeing here is the number of contacts
Neat, right?

What if we want to have a more convenient way to plot this data? Well first we need to give our hi five data a special class for autoplot to work on.

## A brief digression into S3 methods

The reason `plot` works on both a `data.frame` and a `lm` object, is because it is actually _detecting_ what class objects are, and then using a different function - in this case, `graphics:::plot.data.frame`, or `plot.lm`. 

```{r}
fit <- lm(n_hi_five ~ ., data = hi_five_df)
plot(fit)
plot(hi_five_df)
```

Another good example is `summary`:

```{r}
summary(fit)
summary(hi_five_df)
```

Under the hood it uses `summary.lm` and `summary.data.frame`

It's a special shortcut, so that all you need to think about is `summary`, and not:

> what special summary function do I need to call to work on a linear model?

or

> what special plot function do I need to call on my new model?

Another great thing is that you can define _your own_ S3 method, which means that these can be expanded by users - we don't need to try to contribute a new method to base R, we can make our own! So let's do that with autoplot.

# Making an autoplot: first, make your own class

Let's keep it classy.

There are more recommended steps to creating your own class in R, and I'd recommend taking a read of the [S3 chapter in Advanced R]() to get a bit more detail, but we need to create our own class, so that we can create our own autoplot method. This should hopefully make a bit more sense soon.

First, start by creating a function that adds a new class. let's call our class "hi_five"

```{r}
new_hi_five <- function(x){
  class(x) <- c("hi_five", class(x))
  x
}
```

Now before we do this, let's first inspect the class of our hi five data frame:

```{r}
class(hi_five_df)
```

We see it has three classes - `tbl_df`, `tbl`, and `data.frame`. The first two are to do with `tibble` (I'm not sure why there are two - if you know, leave a comment below!), and the last one is it being a `data.frame`

Now we need to use our `new_hi_five` on our data

```{r}
new_hi_five(hi_five_df)
```

... but we need to save it to a new object

```{r}
class_hi_five_df <- new_hi_five(hi_five_df)

class(class_hi_five_df)
```

tada!

Now to help automate that, let's add that line of code to our hi five creation function. (This is why it's fun and cool to write functions - we can modify and update them!)

```{r}
hi_fives_to_df <- function(df){
  df %>% 
  as_tibble(rownames = "age_to") %>% 
  pivot_longer(
    cols = -age_to,
    names_to = "age_from",
    values_to = "n_hi_five"
  ) %>% 
    new_hi_five()
}
```

Put it together

```{r}
hi_five_df <- make_symmetric_hi_fives() %>% 
  hi_fives_to_df()

class(hi_five_df)
```

Woo! Now to define a new S3 method for autoplot, you need to do the following:

```{r}
autoplot.hi_five <- function(object, ...){
  ggplot(object,
       aes(x = age_to,
           y = age_from,
           fill = n_hi_five)) + 
  geom_raster() +
  scale_fill_continuous_sequential("Purples")
}
```

Importantly - you need to follow the same arguments as the original autoplot - which you can check with `?ggplot2::autoplot` - which you'll see is:

```
autoplot(object, ...)
```

Now, we can just use `autoplot` like so!

```{r}
autoplot(hi_five_df)
```

Tada!

Now, you could also make a `matrix` method for this if you wanted as well - say if you just wanted input like this

```{r}
hi_five_mat <- make_symmetric_hi_fives()
hi_five_mat
class(hi_five_mat)
```

Let's say we want to call this class a `hi_five_matrix`, then we can add a new class like this:

```{r}
new_hi_five_matrix <- function(x){
  class(x) <- c("hi_five_matrix", class(x))
  x
}
```


```{r}
hi_five_mat <- make_symmetric_hi_fives() %>% new_hi_five_matrix()
hi_five_mat
```

(Notice those pesky `attr(,"class")` things hanging out down the bottom there, we'll discuss how to fix that in a future blog post, but that is to do with writing custom print methods, just so you know.)

```{r}
autoplot.hi_five_matrix <- function(object, ...){
  data <- hi_fives_to_df(object) %>% 
    mutate(n_hi_five = as.numeric(n_hi_five))
  ggplot(data,
       aes(x = age_to,
           y = age_from,
           fill = n_hi_five)) + 
  geom_raster() +
  scale_fill_continuous_sequential("Purples")
}
```

Now, this is basically the same thing as before - the only different is that you can provide a matrix as the input now.

```{r}
autoplot(hi_five_mat)
```

# How do I do this in an R package?

If you want to use an `autoplot` method in your package, you will need to export the `autoplot` method from `ggplot2`. So you'll need to have some code like this:

```r
#' @importFrom ggplot2 autoplot
#' @export
ggplot2::autoplot
```

Which I usually put in a `pkgname-package.R` file.
